\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-14 ace-pldi14/ace-pldi14.tex
happy thru sec 2.3 metaprogramming

The \emph{target} of compilation is also user-defined. We will show examples of Ace targeting Python as well as OpenCL and CUDA, lower-level languages often used to program graphics hardware. An active base or type can support multiple target languages. Ace functions targeting a language with Python bindings can be called directly from Python scripts, with compilation occurring implicitly. In some cases, types can propagate from Python into Ace. We show how this can be used to streamline the kinds of interactive workflows that Python is often used for. Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external tools.

The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example library that internalizes and extends the OpenCL language. Then in Sec. \ref{att}, we show how this and other libraries are implemented by detailing the extension mechanisms within Ace. To explain and demonstrate the expressiveness of these mechanisms (in particular, active types) further, we continue in Sec.  \ref{examples} by showing a diverse collection of abstractions drawn from different language paradigms that can be implemented as orthogonal libraries in Ace. We include functional datatypes, objects, macros, and typesafe format strings and regular expressions. In Sec.  \ref{related}, we compare Ace to related work on language extensibility and metaprogramming. We conclude in Sec. \ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and discuss limitations and potential future work.

@@

The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax. As a consequence, users of Ace  immediately benefit from an ecosystem of well-developed tools that work with Python code, including parsers, code highlighters, editor modes, style checkers and documentation generators.

\subsection{OpenCL as an Active Library}
The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly. Ace itself provides no built-in support for OpenCL.

To briefly review, in OpenCL, developers can define functions, called {\em kernels}, that execute on \emph{devices} like GPUs or multi-core CPUs \cite{opencl11}. Code on the host CPU launches a kernel over tens of thousands of threads. Each thread executes the same kernel but has access to a unique index, called its \emph{global id} (a data-parallel SPMD programming model). Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce as needed in our examples below.

\subsection{Generic Functions}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
\label{map}
\end{codelisting}
Lines 3 and 4 introduce the definition of  \verb|map|, an Ace function taking three arguments, \verb|input|, \verb|output| and \verb|f|, governed by the active base \verb|clx.std_base|. The body of this function, highlighted in grey for emphasis, does not have Python's semantics. Instead, it will be governed by the active base as well as logic associated with the types of the arguments and expressions it contains.

Generic functions in Ace are instances of \verb|ace.GenericFn|. Generic functions consist of an abstract syntax tree, an active base and an active target\todo{call it this in intro}. The job of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the specified active base and active target. Because no active target was provided, the active base  provides a default target in this case. The default target provided by \verb|clx.std_base| is \verb|clx.opencl|, but, as we will discuss in Sec. \ref{bases}, other bases, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries. One base can support multiple targets.

Python decorators are simply syntactic sugar for applying a function (here, the result of \verb|ace.fn(clx.std_base)|) directly to the function definition being decorated \cite{python-decorators}. In other words, line 3 could be removed and replaced by inserting the equivalent statement \verb|map = ace.fn(clx.std_base)(map)| to line 9. 
 The decorator applied here extracts the abstract syntax tree from \verb|map| using the Python standard  library packages  \verb|inspect| (to retrieve source code) and \verb|ast| (to parse it into a syntax tree).

\subsection{Concrete Functions}
The definition of \verb|map| in Listing \ref{map} could be meaningful for many different type assignments to \verb|input|, \verb|output| and \verb|f|. The body of the function imposes certain constraints on these types, but there is not necessarily a unique type assignment that satisfies them. In a sense, then, \verb|map| is a \emph{family} of functions. Any type assignment for which the operations in the function's body are well-defined can be generated from \verb|map|. We call an Ace function for which types have been assigned a \emph{concrete function}. Concrete functions are instances of \verb|ace.ConcreteFn|. We will show different ways types can be assigned to the arguments of a generic function, and how this, in tandem with the active base, controls the semantics and compilation of the function body in Secs. \ref{procedural} and \ref{implicit}. First, we must understand how types are defined and used in Ace.

\subsection{Types}
In Ace, types are themselves values that can be manipulated at compile-time (unlike most contemporary languages, where user-defined types are simply declared at compile-time and cannot be defined or manipulated programmatically). In particular, types are instances of a Python class that implements the \verb|ace.ActiveType| interface. Python, then, serves also as the \emph{type-level language} of Ace (see related work in Sec. \ref{related}).

Listing 2 shows three types, on lines 7-9, being assigned to variables for convenience (this is similar to \verb|typedef| in C-like languages or \verb|type| in Haskell). These types all come from the \verb|clx| library imported on line 1, none are built into Ace itself.

@@

This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic 
\subsection{Programmatic Compilation}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-5}}]{listing4.py}
\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add\_5} function over two  types of input.}
\label{mapadd5dbl}
\end{codelisting}

\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing2.py`
Hello, compile-time world!
[acec] listing2.cl successfully generated.
\end{lstlisting}
\caption{Compiling \texttt{listing2.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
As defined on Lines 5-7, \verb|main| is a \emph{generic function}. This terminology in Ace is used to indicate a function for which types have not yet been assigned to arguments. Invoking the \verb|compile| method on a generic function with a sequence of argument types invokes the active typechecking and translation mechanism we will describe in Section 3 to produce a \emph{concrete function} -- one with a single type for each argument, internal variable (such as \verb|hello|) and the return value. In Listing \ref{hello}, we name this concrete function \verb|main|, overwriting the generic function of the same name that it is derived from. The compiler does nothing apparently interesting: there are no arguments, the single internal variable takes the OpenCL string type \verb|char*| from its value and the return type is \verb|void|.

@@

\subsection{Whole-Function Type Inference}
\begin{codelisting}[t]
\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
\label{metaprogramming}
\end{codelisting}
\begin{codelisting}
\lstinputlisting{listing8.py}
\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
\label{py}
\end{codelisting}

%\begin{codelisting}
\lstinputlisting{listing6.py}
%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
%\label{inference}
%\end{codelisting}

@@

\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}
Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}

@@

\subsection{Metaprogramming in Ace}
Metaprogramming refers to the practice of writing programs that manipulate other programs. There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{klockner2011pycuda}. OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming. Ace supports programmatic compilation, higher-order functions and a flexible language extension mechanism, so many use cases for metaprogramming have been eliminated.

However, there are still valid uses for this feature and Ace fully supports it. On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_str| variant of the \verb|OpenCL.fn| method. It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10. The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|. This transformation as well as some others are distributed in \verb|ace.astx| for convenience.

\section{Extensibility}\label{att}
Thus far, we have been discussing the \verb|OpenCL| module in our examples. This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators \cite{opencl11}. The functions we wrote have used these primitives with the \verb|OpenCL.OpenCL| backend, so the translation has been direct and no run-time overhead has been introduced. Thus, as described so far, Ace has affirmatively resolved the question it was originally conceived to address, discussed in Section 2.1.

>>>>

The \emph{target} of compilation is also user-defined. We will show examples of Ace targeting Python as well as OpenCL and CUDA, lower-level languages often used to program graphics hardware. An active base or type can support multiple \emph{active targets}, which mediate translation of Ace code to code in a target language. Ace functions targeting a language with Python bindings can be called directly from Python scripts, with compilation occurring implicitly. For some data structures, types can propagate from Python into Ace. We show how this can be used to streamline the kinds of interactive workflows that Python is often used for. Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external means.

The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example library that internalizes and extends the OpenCL language. Then in Sec. \ref{att}, we show how this and other libraries are implemented by detailing the extension mechanisms within Ace. To explain and demonstrate the expressiveness of these mechanisms (in particular, active types) further, we continue in Sec.  \ref{examples} by showing a diverse collection of abstractions drawn from different language paradigms that can be implemented as orthogonal libraries in Ace. We include functional datatypes, objects, macros, and typesafe format strings and regular expressions. In Sec.  \ref{related}, we compare Ace to related work on language extensibility and metaprogramming. We conclude in Sec. \ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and describing their limitations and potential future work.

@@

The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax. As a consequence, users of Ace  immediately benefit from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.

\subsection{OpenCL as an Active Library}
The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly. Ace itself has no built-in support for OpenCL.

To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}. Each thread executes the same kernel but has access to a unique index, called its \emph{global ID}. Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce  as needed in our examples below.

\subsection{Generic Functions}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
\label{map}
\end{codelisting}
\begin{codelisting}[t]
\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
\label{metaprogramming}
\end{codelisting}
Lines 3-4 introduce \verb|map|, an Ace function governed by the active base \verb|clx.base| taking three arguments. The \emph{active target}, \verb|clx.opencl|, determines which language the function will target (here, the OpenCL kernel language) and mediates code generation.

The body of this function, highlighted in grey for emphasis, does not have Python's semantics. Instead, it will be governed by the active base together with the active types used within it. No such types have been provided explicitly for the arguments yet, however. Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments, so \verb|map| is not yet a \emph{concrete function}. We call functions awaiting types \emph{generic functions}. %Once types have been assigned, they are called \emph{concrete functions}.

Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}. The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target. 
Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}. In other words, line 3 could be replaced by inserting the following  statement on line 9:

\verb|map = ace.fn(clx.base, clx.opencl)(map)|

The abstract syntax tree for \verb|map| is extracted using the Python standard  library packages  \verb|inspect| (to retrieve its source code) and \verb|ast| (to parse it into a syntax tree).

\subsection{Metaprogramming in Ace}
Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming. Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|thresh_scale| and \verb|thresh_dbl|. The latter is derived from the former using a library for manipulating Python syntax trees, \verb|astx|. In particular, its \verb|specialize| function replaces uses of the second argument of \verb|thresh_scale| with the literal \verb|2.0| (and changes the function's name), leaving a function of one argument.

\subsection{Types}
To go from a generic function to a \emph{concrete function}, we must provide a type assignment for its arguments. We will show different ways to do this in Secs. \ref{procedural} and \ref{implicit}. Before we can do this, however, we must understand what types are.

In Ace, types are themselves values that can be manipulated at compile-time (unlike most contemporary languages, where user-defined types are simply declared at compile-time and cannot be defined or manipulated programmatically). In particular, types are instances of a Python class that implements the \verb|ace.ActiveType| interface. Python, then, serves also as the \emph{type-level language} of Ace (see related work in Sec. \ref{related}).

Listing 2 shows three types, on lines 7-9, being assigned to variables for convenience (a construct similar to \verb|typedef| in C-like languages or \verb|type| in Haskell). These types all come from the \verb|clx| library imported on line 1, none are built into Ace itself.

@@

This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic 
\subsection{Programmatic Compilation}\label{procedural}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-5}}]{listing4.py}
\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add\_5} function over two  types of input.}
\label{mapadd5dbl}
\end{codelisting}

\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing3.py`
Hello, compile-time world!
[acec] listing3.cl successfully generated.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{mapadd5dbl}.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
As defined on Lines 5-7, \verb|main| is a \emph{generic function}. This terminology in Ace is used to indicate a function for which types have not yet been assigned to arguments. Invoking the \verb|compile| method on a generic function with a sequence of argument types invokes the active typechecking and translation mechanism we will describe in Section 3 to produce a \emph{concrete function} -- one with a single type for each argument, internal variable (such as \verb|hello|) and the return value. In Listing \ref{hello}, we name this concrete function \verb|main|, overwriting the generic function of the same name that it is derived from. The compiler does nothing apparently interesting: there are no arguments, the single internal variable takes the OpenCL string type \verb|char*| from its value and the return type is \verb|void|.

@@

\subsection{Whole-Function Type Inference}
\begin{codelisting}
\lstinputlisting{listing8.py}
\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
\label{py}
\end{codelisting}

%\begin{codelisting}
%\lstinputlisting{listing6.py}
%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
%\label{inference}
%\end{codelisting}

@@

\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}

@@

\section{Extensibility}\label{att}
If no active target is explicitly provided, as in our example, the active base  provides a default target. The default target provided by \verb|clx.base| is \verb|clx.opencl|, but other targets, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries. One base can support multiple targets. We will discuss targets in Sec. \ref{targets}.

Thus far, we have been discussing the \verb|OpenCL| module in our examples. This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators \cite{opencl11}. The functions we wrote have used these primitives with the \verb|OpenCL.OpenCL| backend, so the translation has been direct and no run-time overhead has been introduced. Thus, as described so far, Ace has affirmatively resolved the question it was originally conceived to address, discussed in Section 2.1.
